### 2017-10-14 CCTValproject/CCTVal-Project.tex
Last approved version

\title[CCTVal Research project]{CCTVal Research project}

@@

The findings of this project could be easily used as an integration tool
to detect arbitrage opportunities or risk control in financial time series. Also, the methodology could be extended to other econometric methods which could be benefited of machine learning techniques.

@@

VECM introduces the long-run
relationship among a set of cointegrated time series as an error correction
term. These time series must be integrated of order 1, denoted I(1), i.e. they
become stationary at their first differences. In finance, I(1) time series are
very common and to introduce cointegration restrictions in models often improves
forecasting, see \cite{duy1998}. Therefore, VECM has been widely adopted in
financial applications \cite{mukherjee1995}, \cite{seong2013},
\cite{maysami2000} and \cite{arestis2001}. VECM has also been used in pair
trading, see \cite{herlemont2003}, or models with more than two variables, see
for example \cite{mukherjee1995} and \cite{engle2004}.

Cointegration relationships can be found in low and high frequency data of two or
more assets. While cointegration in low frequency data is motivated by a
long-run equilibrium relationship between economic forces, cointegration in high
frequency data has its foundation in statistical arbitrage theory which is very helpful
to detect mean-reverting trades. Information about cointegrated assets in high
frequency data could be used as an input for high frequency trading strategies,
using it as a signal to capture small profits in short term trades, see
\cite{miao2014}. \cite{zhou2001} addressed the benefits of using higher
frequency data to analize cointegration. \cite{rittler2012} also found that
cointegration may differ with different data frequencies and could appear
with increased data frequency.

The use of VECM with high frequency data is mainly limited by computationally
expensive routines. Firstly, VECM parameters are obtained using the ordinary
least squares (OLS) it has over-fitting
issues and is computationally expensive when
the number of lagged values and data increases. This problem is commonly adressed using
Ridge Regression (RR) \cite{hoerl1970} which introduces a regularisation
parameter that leads to an unbiased estimation with better generalisation
capability. Secondly, obtaining
cointegration vectors is also an expensive routine because the Johansen method 
is of order $O(n^3)$ (\cite{johansen1995}).
\cite{chen2003}, \cite{Arce2017} addressed the advantage of distributed processing over
conventional rolling window processing. Therefore, our aim is to design
an online method based on cointegration using RR instead of OLS and that works on a distributed environment so it can be used with high frequency stream data.

\section{Research Design and Methods}
We intend to extend the approach presented in \cite{icpram15}, where the proposed method called OVECM only considered a time varying window of historical data using OLS to
get model parameters. This proposal is about using RR or its variant  The  Aggregating Algorithm for Regression (AAR)  \cite{vovk2001} as an alternative of OLS to estimate model parameters.

Since cointegration vectors represent a long-run relationship between the time
series, they vary little in time. This proposal also includes to obtain new
cointegration vectors only when this long-term relationship changes.
We will estimate the performance of our method using different frequencies of financial time series, so we can measure the effect of cointegration in high frequency data.

@@

We also hypothesise that, using high frequency data, cointegration vectors will remain the same for some period of time, since they represent long-run relationship.

>>>>

\title[CCTVal Research project proposal]{CCTVal Research project proposal}

@@

The findings of this project could be easily used as an integration tool
to detect arbitrage opportunities or risk control in financial time series. Also, the methodology could be extended to other econometric methods which could be enhanced by machine learning techniques.

@@

In order to use the VECM, the time series must be integrated of order 1, denoted I(1), i.e. they
become stationary at their first differences. In finance, I(1) time series are
very common. VECM introduces the cointegration relationship as an error correction
term which often improves forecasting, see \cite{duy1998}. Therefore, VECM has been widely adopted in
financial applications \cite{mukherjee1995,seong2013,maysami2000,arestis2001}. VECM has also been used in pair trading, see \cite{herlemont2003}, or models with more than two variables, see
for example \cite{mukherjee1995} and \cite{engle2004}.

Cointegration relationships can be found in low and high frequency data of two or
more assets. While cointegration in low frequency data is motivated by a
long-run equilibrium relationship between economic forces, cointegration in high
frequency data has its foundation in statistical arbitrage theory which is very helpful
to detect mean-reverting trades. Information about cointegrated assets in high
frequency data could be used as an input for high frequency trading strategies,
using it as a signal to capture small profits in short term trades, see
\cite{miao2014}. \cite{zhou2001} addressed the benefits of using higher
frequency data to analyse cointegration. \cite{rittler2012} also found that
cointegration may differ with different data frequencies and could appear
with increased data frequency.

The use of VECM with high frequency data is mainly limited by computationally
expensive routines:
\begin{itemize}
\item VECM parameters are obtained using the ordinary
least squares (OLS) which has over-fitting
issues and is computationally expensive when
the number of lagged values and data increases. This problem is commonly addressed using
Ridge Regression (RR) \cite{hoerl1970}. RR introduces a regularisation
parameter that leads to an unbiased estimation with better generalisation
capability. 
\item The Johansen method is used to obtain 
cointegration vectors and it is also an expensive routine  of order $O(n^3)$  \cite{johansen1995}. A distributed processing approach has been used recently successfully reducing execution times, see \cite{chen2003}, \cite{Arce2017}. Specially in \cite{Arce2017}we showed the advantage of distributed processing over conventional rolling window processing.
\end{itemize}
Therefore, our aim is to design an online method based on cointegration using RR instead of OLS and that works on a distributed environment so it can be used with high frequency stream data.

\section{Research Design and Methods}

In this proposal, we will extend the approach presented in \cite{icpram15}. In this article we proposed method called OVECM that only considered a time varying window of historical data using OLS to get model parameters. This proposal will explore the use of RR or its variant The  Aggregating Algorithm for Regression (AAR)  \cite{vovk2001} as an alternative of OLS to estimate model parameters.

Additionally, since cointegration vectors represent a long-run relationship between the time
series, they vary little in time. Therefore, we plan to obtain new
cointegration vectors only when this long-term relationship changes, this can be tracked using a threshold of forecast error for example.

We will estimate the performance of our method using different frequencies of financial time series, so we can measure the effect of cointegration in high frequency data.

@@

We also hypothesise that, using high frequency data, cointegration vectors will remain the same for some period of time, since they represent long-run relationships.

### 2017-10-14 src/1_Introduction.tex
Last approved version

"An individual economic variable, views as a time series, can wander extensively
and yet some pairs of series may be expected to move so that they do not drift
far apart."-Robert F. Engle and Clive W.J. Granger~\cite{engle1987}.\\
In this introduction we present the scope, objectives, hypothesis and
organization of this thesis.

@@

The specific objectives of this research are,
\begin{itemize}
\item $\Large \mathcal{O}_1$: \emph{A review of the literature on time series
analysis models including machine learning techniques.}
\item $\Large \mathcal{O}_2$: \emph{Development of a set of known features of the
studied time series and applied them to improve forecasting.}
\item $\Large \mathcal{O}_3$: \emph{Development of parallel and efficient
algorithms to ensure quick response times .}
\item $\Large \mathcal{O}_4$: \emph{Deep mathematical analysis of the proposal
and financial concepts involved.}
\item $\Large \mathcal{O}_5$: \emph{Design and implement representative set of
experiments in order to show when and why proposal performs better.}
\end{itemize}

>>>>

"An individual economic variable, viewed as a time series, can wander extensively
and yet some pairs of series may be expected to move so that they do not drift
far apart."-Robert F. Engle and Clive W.J. Granger~\cite{engle1987}.\\
In this introduction we present the scope, objectives, hypothesis and
organization of this thesis.

@@

The specific objectives of this research are,
\begin{itemize}
\item $\Large \mathcal{O}_1$: \emph{A review of the literature on time series
analysis models including machine learning techniques.}
\item $\Large \mathcal{O}_2$: \emph{Development of a set of known features of the
studied time series and the application to improve forecasting.}
\item $\Large \mathcal{O}_3$: \emph{Development of parallel and efficient
algorithms to ensure quick response times .}
\item $\Large \mathcal{O}_4$: \emph{Deep mathematical analysis of the proposal
and financial concepts involved.}
\item $\Large \mathcal{O}_5$: \emph{Design and implement representative set of
experiments in order to show when and why the proposal performs better.}
\end{itemize}

### 2017-10-14 src/5_1_Proposal.tex
Last approved version

The parallel implementation allowed the execution times to be reduced more than
9 times and therefore a response time was obtain before 10 seconds. Since we
used 10-second frequencys we can say that our proposal is suitable for use in an
online context for real applications because response times were less than this
frequency. Cointegration information can now easily be used as an integration tool
to detect arbitrage opportunities or risk control.

>>>>

The parallel implementation allowed the execution times to be reduced more than
9 times and therefore a response time was obtain before 10 seconds. Since we
used 10-second frequencies we can say that our proposal is suitable for use in an
online context for real applications because response times were less than this
frequency. Cointegration information can now easily be used as an integration tool
to detect arbitrage opportunities or risk control.

### 2017-10-14 src/6_Conclusion.tex
Solo falta relacion con los objetivos en capitulo de conclusiones
Last approved version

\chapter{Conclusion}

@@

@@

@@

>>>>

\chapter{Conclusions and Future Work}
\label{chapter:conclusions}
Main results and contributions are presented in this chapter along with some
suggested future work.

\vspace{0.5cm}

@@

This thesis is a multidisciplinary work that involves knowledge of Finance
and Economics, Time Series, Machine Learning, Parallel Computing and Scientific
Computing. We specifically addressed the joint dynamic behaviour of financial time
series that are said to be cointegrated. A broadly used approach, called VECM (Vector
Error Correction Model), characterises this behaviour in terms of forces pulling
towards equilibrium called cointegration. However, the use of VECM has been
limited to low frequency time series processed in batch mode. In this thesis we
explored the use of VECM with high frequency data and we found that the main
limitation was computational. The focus of our work was developing a model to improve the performance forecasting of financial time series maintaining good execution times in order to use it with high frequency data. VECM parameter estimation uses two computationally
expensive routines: the Johansen method, to obtain cointegration vectors, and
the ordinary least squares method to solve the system. In this thesis, different
ways to explore cointegration in high frequency time series were proposed,
always considering the computational limitations of the VECM.

The study of cointegration in high frequency data was done using two different
approaches presented in Chapters \ref{chapter:proposal1} and
\ref{chapter:proposal2}, called AVECM and OVECM:
\begin{description}
\item[AVECM] AVECM is an adaptive version of VECM which includes a new method to choose VECM parameters based on
the maximisation of the percentage of cointegration. AVECM was implemented using
MPI to search on a grid of possible values. We evaluated our results in terms of performance and execution times:
\begin{description}
\item[Performance] Results showed that AVECM improves
performance measures by finding parameters of L and p maximising the percentage
of cointegration. AVECM performance was compared against ARIMA and the random walk model. We showed that the out-of-sample performance for AVECM is superior to ARIMA and the naive random walk model in terms of the MSE and $U$-statistic. This result is very important since it is still difficult to outperform ARIMA and the random walk model for standard econometric forecasting models  despite their simplicity. 
\item[Execution times] execution times were reduced more than 9 times
ensuring a response time before the processing of the next data point. Despite the fact that high frequency Forex data can be spurious, the model
performance can be less reliable (and more spurious) relative to the lower
frequencies (such as 1 minute or 5 minute intervals) adopted by some other
studies. However, the deficiency is offset by gain in accuracy from parallel
processing which is capable of searching or examining a much larger state space
given the same computational time.
\end{description}
\item[OVECM] We proposed an online version of VECM (OVECM). OVECM optimises how
model parameters are obtained using a sliding window of the most recent data.
OVECM, unlike AVECM, updates the parameters at each step, instead of obtaining
new ones. This proposal takes advantage of the long-run relationship
between the time series in order to obtain improved execution times. OVECM also
introduces matrix optimisation in order to obtain the new model in an iterative
way.  
\begin{description}
\item[Performance] 
OVECM was compared with the traditional VECM and ARIMA. Despite the fact that traditional VECM slightly outperformed our proposal, the OVECM execution time is lower. This result can be explained because OVECM avoids the calculation of cointegration vectors at each time step which improves execution times but can affect forecast performance.
\item[Execution times] OVECM was compared with the traditional VECM and ARIMA
with the same sliding window sizes. As a result, OVECM outperformed both in terms of
execution time. This reduction of execution time is mainly because OVECM avoids the cointegration vector calculation using the Johansen method. Aditionally, OVECM took much less than a minute at every
step making it possible to use with higher frequency data.  
\end{description}
\end{description}
Finally, cointegration information can now easily be used as an integration tool
to detect arbitrage opportunities or risk control in financial time series.

@@

This research helped to integrate machine learning techniques and parallel computing with econometric models such as VECM so they can be used with high frequency data. We also observed that cointegration relationships are sensitive to the choice of the amount of data and VECM parameters and this can be used to improve forecasting performance. We proposed two variations of VECM, OVECM and AVECM which could be extended to other econometric models.

Regarding the initial objectives defined at the beginning of this thesis we can say:

\begin{itemize}
\item $\Large \mathcal{O}_1$: \emph{A review of the literature on time series
analysis models including machine learning techniques.}\\
The literature review was focused in finance, time series concepts and models and machine learning including theory and applications.
\item $\Large \mathcal{O}_2$: \emph{Development of a set of known features of the
studied time series and the application to improve forecasting.} \\
We included specific knowledge about forex rates, which were those used in our experiments, such as trade best trading times, stylised facts, percentage of cointegration, time series frequency, among others.
\item $\Large \mathcal{O}_3$: \emph{Development of parallel and efficient
algorithms to ensure quick response times .} \\
We showed that high performance computing will allow the use of computationally expensive methods to forecast high frequency data ensuring quick response times.
\item $\Large \mathcal{O}_4$: \emph{Deep mathematical analysis of the proposal
and financial concepts involved.} \\
The two proposals were presented with a strong mathematical foundation, we added definitions and proofs of all the concepts involved.
\item $\Large \mathcal{O}_5$: \emph{Design and implement representative set of
experiments in order to show when and why the proposal performs better.}
All the experiments were designed to give a general view of the method performance. We clearly showed the use and limitations of our proposals.
\end{itemize}

@@

For future study, it would be interesting to explore the relationship between
cointegration and performance for more assets, including not only forex rates
but also stocks. It would also be interesting to include more explaining
variables such as bid-ask spread and change in volume.

The online approach for other econometrics models it will be also worthy of 
study. Many of them could be adapted and used with higher frequency data. In
this thesis, the online version of OLS was studied, but the online version of
ridge regression could also be applied to obtain a solution with better
generalisation capabilities.

Finally, it will be also interesting to improve the out-of-sample forecast
by considering more explicative variables, to increase window sizes or try different
conditions to obtain new cointegration vectors.
